---
title: "Genetics of Evolution - Mechanisms"
engine: knitr
format: 
  html: 
    toc: true
    toc-depth: 3
    toc-title: Contents
    self-contained: true
    css: ../custom.css
    link-external-newwindow: true
webr: 
  packages: ["ggplot2", "reshape", "tibble", "viridis", "dplyr"]
filters: 
  - webr
bibliography: ../Genetics_Discussion.bib
csl: ../evolution.csl
code-annotations: hover
---


```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(tibble)
library(tidyr)
library(reshape)
library(viridis)
library(dplyr)

ggplot2::theme_set(ggplot2::theme_light())

source("drift.R")
```


```{webr-r}
#| context: setup

library(ggplot2)
library(tibble)
library(tidyr)
library(reshape)
library(viridis)
library(dplyr)

ggplot2::theme_set(ggplot2::theme_light())

source("https://raw.githubusercontent.com/BIOSC-2200/BIOSC-2200.github.io/main/GE/drift.R")

```


## Preliminaries

If you are not already familiar with the structure of these exercises, read the [Introduction](https://biosc-2200.github.io/Introduction/) first.

::: {.callout-note}

Reminder: Save your work regularly.

:::

::: {.callout-important}

If you are using a Mac, we recommend that you use either Chrome or Firefox to complete these exercises. Some of the default settings in Safari prevent these exercises from running.

:::


### Contact information

If you have questions about these exercises, please contact Dr. Kevin Middleton (middletonk@missouri.edu) or drop by Tucker 224.


## Learning objectives

The learning objectives for this exercise are:

- Describe and identify the mechanisms by which variation arises and is fixed (or lost) in a population over time.
- Explain how the processes of drift, natural selection, migration, and mutation can affect the elimination, maintenance or increase in frequency of alleles in a population.




## Exploring the assumptions of HWE


Mechanism of selection

	- Moving population average
	- Drift vs. selection
- Effect of sample size on selection and drift in phenotypic evolution
	- Drift as sampling error
	- Drift has a larger effect in small populations

Populations have significant underlying variation

Intuition of moving mean


```{r}
#| echo: false
n_Populations <- 10
n_Generations <- 100

Population_Size <- 100
Initital_Frequency <- 0.5

Fitness_AA <- 1
Fitness_AB <- 1
Fitness_BB <- 1

Migration <- 0

Mutation_AB <- 0
Mutation_BA <- 0

Simulate_Population()
```


```{webr-r}
n_Populations <- 10
n_Generations <- 100
n <- 100
p <- 0.5

Fitness_AA <- 1
Fitness_AB <- 1
Fitness_BB <- 1

Migration <- 0

Mutation_AB <- 0
Mutation_BA <- 0

Simulate_Population()
```


## Feedback FIXME

We would appreciate your anonymous feedback on this exercise. If you choose to, [please fill out this optional 4-question survey](https://missouri.qualtrics.com/jfe/form/SV_25nuDUfHEl1cQku) to help us improve.


## References

::: {#refs}
:::
