---
title: "Case study: Investigating a newly discovered muscle mutation in mice"
engine: knitr
format: 
  html: 
    toc: true
    toc-depth: 3
    toc-title: Contents
    self-contained: true
    css: ../custom.css
    link-external-newwindow: true
  docx:
    reference-doc: word-styles-reference-01.docx
webr: 
  packages: ["ggplot2", "dplyr", "forcats", "tibble"]
filters: 
  - webr
bibliography: ../Genetics_Discussion.bib
csl: ../evolution.csl
code-annotations: hover
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(tidyr)
library(dplyr)
library(gt)
library(grid)
library(patchwork)
library(magick)

ggplot2::theme_set(ggplot2::theme_light())
```

```{webr-r}
#| context: setup
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(forcats)
library(tibble)

ggplot2::theme_set(ggplot2::theme_light())
```


## Preliminaries

If you have not already, start with the first part of this exercise: [Transmission of Genetic Information](https://biosc-2200.github.io/TGI/). That page has the background and information you need to fully understand this case study.


## Introduction

In the early 1990's, Dr. Theodore Garland and colleagues began a [long-term experiment](https://sites.google.com/ucr.edu/hrmice/home) breeding experiment using house mice (*Mus musculus*). The goal of this experiment was to explore the inter-relatedness of traits involved in physiology, anatomy, brain function, immunology, behavior, and many others. 

From a random, genetically diverse initial population, the research team established eight separate groups of mice. Four of the groups would be artificially selected for high levels of voluntary activity (measured using a running wheel attached to the cage) and the other four would be bred randomly. [Here is a link](https://biology.ucr.edu/people/faculty/Garland/Herron_and_Freeman_2014_pages_39-40_Garland.pdf) to a book section that describes the experiment. [Here is a movie of what the mice look like on the running wheels.](https://www.youtube.com/watch?v=RuqhC7g_XP0).

Because wheel-activity, the trait being selected, is heritable, the two sets of lines rapidly diverged (@fig-run).

![Total wheel revolutions per day for 4 selected and 4 control lines of mice across the first 10 generations of selection. From Swallow et al. [-@Swallow1998-xo]](../images/garland1.png){width=75% fig-align="center" fig-alt="" #fig-run}

Among other phenotypes, Garland and colleagues were interested in studying muscle physiology. Did the selected mice evolve to have different muscle masses or muscle fiber types? They were routinely weighing the calf muscles of mice as part of their research and generated plots that looked like this:

```{r}
#| echo: false
#| fig-height: 3
#| fig-align: center

MM <- readr::read_rds("../data/MouseMuscle.rds") |> 
  mutate(Line = factor(Line)) |> 
  drop_na()

ggplot(MM, aes(x = Body_Mass, y = Muscle_Mass, color = Line)) +
  facet_grid(. ~ Sex) +
  geom_point(size = 3) +
  scale_color_manual(values = c("aquamarine4", "navy", "firebrick4")) +
  labs(x = "Body Mass (g)", y = "Muscle Mass (g)") +
  theme(strip.background = NULL,
        strip.text = element_text(color = "black", face = "bold",
                                  size = 12),
        axis.title = element_text(face = "bold"))
```

This plot shows the mass of the calf muscles plotted against body mass. In general you expect to find that larger mice will have larger muscles, just like an adult human has larger muscles than a child. But you do not expect that one of the lines of mice would have such disproportionately small muscles compared to the others. Generation after generation, they found that some of the mice had muscles that weighed only about 50% as much as they would be expected to based on their body mass. In other words, there appeared to be a subset of mice that had muscles which were only half as large as predicted.

They found this small muscle phenotype in three different lines, and estimated the prevalence across generations (@fig-mmgen). These authors also found that the small muscle phenotype was favored by selection in these mice [@Houle-Leroy2003-tm].

![Proportion of mice exhibiting the small muscle phenotype in three lines. The squares in this plot (Line 3) are the same line as the blue points in the plot above. From Garland et al. [-@Garland2002-it]](../images/garland2.png){width=75% fig-align="center" fig-alt="" #fig-mmgen}

### Developing the hypothesis

Garland and colleagues examined pedigrees of mice that showed the small muscle phenotype. They found patterns like the ones below. In the pedigrees, hatching indicates that the mouse has the small muscle phenotype.


```{r}
#| echo: false
#| fig-height: 3

library(pedtools)

set.seed(32474)

pd1 <- nuclearPed(nch = 10, father = "Male",
                 mother = "Female", 
                 sex = sample(1:2, size = 10, replace = TRUE),
                 children = 1:10)

plot(pd1, hatched = c("Female", "Male", as.character(1:10)))
```

```{r}
#| echo: false
#| fig-height: 3

pd2 <- nuclearPed(nch = 10, father = "Male",
                 mother = "Female", 
                 sex = sample(1:2, size = 8, replace = TRUE),
                 children = 1:8)

plot(pd2, hatched = as.character(c(1, 3, 4, 8)))
```


> What kind of inheritance pattern do you think could explain the observed distribution of phenotypes (dominant vs. recessive, sex-linked vs. autosomal)? Briefly explain your reasoning.

<textarea rows="3" cols="75">

</textarea>


### Testing the hypothesis

Garland and colleagues hypothesized that the small muscle phenotype was caused by an autosomal recessive allele. They set out to determine if their hypothesis was correct by designing a test cross.

The original strain of mice that had been used to produce the 8 lines was outbred, meaning that it had extensive genetic variability.^[The specific strain *ICR* was chosen because it had a similar level of genetic variation as found in humans. If there were no genetic variation, then selective breeding would not have worked.] For the test cross, they chose to use a standard inbred strain of mouse: C57/Bl6 ("C57 Black 6"). This mouse did not show the small muscle phenotype.

For the other parental line, they chose all mice that had the small muscle phenotype.

> Using a piece of paper, set up a Punnett square with a homozygous dominant C57/Bl6 father (*MM* genotype) and a homozygous recessive small muscle phenotype mother (*mm* genotype). What genotypes and phenotypes do you predict if the hypothesis is correct? Describe your results below.

<textarea rows="3" cols="75">

</textarea>

The F~1~ generation should all be *Mm* (all combinations of alleles will be *Mm*). The resulting muscle phenotypes will be normal if the allele is recessive (a single copy of *M* is enough to result in the normal phenotype).

Garland and colleagues found the expected phenotypic pattern (but still didn't know the underlying genotypes). Next they set up a second cross between the F~1~ mice resulting from the first cross and the small muscle phenotype parental female mice (this is called a "backcross").

> Using a piece of paper, set up a Punnett square with a heterozygous father (*Mm* genotype) and a homozygous recessive small muscle phenotype mother (*mm* genotype). What genotypes and phenotypes do you predict if the hypothesis is correct? Describe your results below.

<textarea rows="3" cols="75">

</textarea>

In this cross, genotypes of half of the mice should be heterozygous (*Mm*) and half homozygous for the small muscle allele (*mm*). The phenotypic ratio will also be 1:1, because the heterozygous mice will have the normal muscle phenotype.


### Results

In total, there were 404 mice in the F~2~ backcross generation [@hannon_phenotypic_2008a]. The authors weighed the calf muscles of all the mice (called "triceps surae" here). The results are shown in @fig-f2.

![Observed muscle masses  compared to body mass in a sample of 404 F~2~ backcross mice. From Hannon et al. [-@hannon_phenotypic_2008a].](../images/garland3.png){width=75% fig-align="center" fig-alt="" #fig-f2}

> What is your prediction for the phenotypic ratio from jsut looking at the plot above? Does it appear that there is a 1:1 ratio of normal to small muscle mice?

<textarea rows="3" cols="75">

</textarea>

Based on the observed muscle masses, 201 mice were classified as having the small muscle phenotype and 203 were classified and normal.

```{r}
#| echo: false

tribble(
  ~ Phenotype, ~ Count,
  "Normal", 203,
  "Small Muscle", 201
) |> 
  gt() |> 
  cols_label(Phenotype = md("**Phenotype**"),
             Count = md("**Count**"))

```


### Testing the results

Now it's your turn. In the code block below, we set up the data just like we did for the date of birth example.

```{webr-r}
MM <- tribble(
  ~ Phenotype, ~ Count,
  "Normal", 203,
  "Small Muscle", 201
)
MM
```

> Based on the observed counts, what do you predict that the test will show? Will the counts differ from a 1:1 ratio or will they not differ (the null hypothesis)?

<textarea rows="3" cols="75">

</textarea>


In the code block below, use the counts above to carry out a chi-squared test for the data. You just need to add the column with the data. Look at the example above in the section "Shortcut" if you need a hint.

```{webr-r}
chisq.test()
```

> What are the $\chi^2$ value, the degrees of freedom, and the associated *P*-value? What does this test tell you about the observed ratio of counts?

<textarea rows="3" cols="75">

</textarea>

> How do you think the results might have been different if the authors had only measured 40 mice instead of ~400? Feel free to change the numbers above and re-run the test. Just be sure to run the code block where the MM object is created before you rerun the chi-squared test.

<textarea rows="3" cols="75">

</textarea>


### Epilogue

The "mini-muscle" allele was confirmed as [autosomal recessive](https://www.genome.gov/genetics-glossary/Autosomal-Recessive-Disorder), just as Garland and colleagues hypothesized. The full description of the experiment and a large amount of data and discussion can be found in the original publication [@hannon_phenotypic_2008a]. 

After some work on the genetics of the small muscle phenotype, it was mapped to a 2.6 megabase region on chromosome 11 that contained about 100 genes [@Hartmann2008-ft]. After several more years of more research, the allele was finally identified as a single nucleotide polymorphism (SNP) located in the Myosin Heavy Chain 4 gene [@Kelly2013-td]. Homozygous mice with two copies of this allele (named *Myh4^minimsc^*) fail to produce one type of myosin, which results in small muscles that produce less contractile force but do not fatigue as quickly as unaffected muscles.


## References

::: {#refs}
:::
